---
title: "In-class Exercise 1"
author: "Lee Chee Tian"
editor: visual
---

## Overview

This exercise guides us on the computation of spatial weights in geospatial analytics.

## Getting Started

The code chunk below will install and load tidyverse, sf, spdep and tmap packages.

```{r}
pacman::p_load(sf, tidyverse, spdep, tmap)
```

## Importing Geospatial Data

### Importing polygon features

This code chunk will import ESRI shapefile into R.

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
```

### Importing and transforming into projected coordinates system

```{r}
preschool = st_read("data/geospatial/pre-schools-location-kml.kml") %>%
  st_transform(crs = 3414)
```

### Importing polygon features for Hunan

```{r}
hunan_sf <- st_read(dsn = "data/geospatial", layer = "Hunan")
```

### Importing attribute data in csv

```{r}
hunan = read_csv("data/aspatial/Hunan_2012.csv") 
```

### Relational join of Hunan data

```{r}
hunan <- left_join(hunan_sf, hunan)
```

## Visualizing Regional Development Indicator

```{r}
basemap <- tm_shape(hunan) +
  tm_polygons() +
  tm_text("NAME_3", size=0.5)

gdppc <- qtm(hunan, "GDPPC")
tmap_arrange(basemap, gdppc, asp=1, ncol=2)
```

## Computing Contiguity Spatial Weights

### Computing (QUEEN) contiguity based neighbours

```{r}
wm_q <- poly2nb(hunan, queen = TRUE)
summary(wm_q)
```

```{r}
neighbours_1 <- wm_q[[1]]
neighbours_1
```

```{r}
hunan$County[1]
```

```{r}
hunan$NAME_3[neighbours_1]
```

```{r}
nb1 <- hunan$GDPPC[neighbours_1]
nb1
```

```{r}
str(wm_q)
```

### Creating (ROOK) contiguity based neighbours

```{r}
wm_r <- poly2nb(hunan, queen = FALSE)
summary(wm_r)
```

### Visualizing contiguity weights

```{r}
longitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])
```

```{r}
latitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])
```

```{r}
coords <- cbind(longitude, latitude)
```

```{r}
head(coords)
```

#### Plotting Queen contiguity based neighbours map

```{r}
plot(hunan$geometry, border = "lightgrey")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

#### Plotting Rook contiguity based neighbours map

```{r}
plot(hunan$geometry, border = "lightgrey")
plot(wm_r, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

#### Plotting both Queen and Rook contiguity based neighbours maps

```{r}
par(mfrow=c(1,2))
plot(hunan$geometry, border = "lightgrey")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col = "red", main = "Queen Contiguity")
plot(hunan$geometry, border = "lightgrey")
plot(wm_r, coords, pch = 19, cex = 0.6, add = TRUE, col = "red", main = "Rook Contiguity")
```

## Computing distance based neighbours

### Determine the cut-off distance

```{r}
#coords <- coordinates(hunan)
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)
```

### Computing fixed distance weight matrix

```{r}
wm_d62 <- dnearneigh(coords, 0, 62, longlat = TRUE)
wm_d62
```

**Quiz answer:**\
The average number of links probably means that on average, each region has this number of neighbours that qualify the specified cut-off distance.

```{r}
str(wm_d62)
```

```{r}
table(hunan$County, card(wm_d62))
```

```{r}
n_comp <- n.comp.nb(wm_d62)
n_comp$nc
```

```{r}
table(n_comp$comp.id)
```

#### Plotting fixed distance weight matrix

```{r}
plot(hunan$geometry, border = "lightgrey")
plot(wm_d62, coords, add = TRUE)
plot(k1, coords, add = TRUE, col = "red", length = 0.08)
```

```{r}
par(mfrow = c(1,2))
plot(hunan$geometry, border = "lightgrey")
plot(k1, coords, add = TRUE, col = "red", length = 0.08, main = "1st nearest neighbours")
plot(hunan$geometry, border = "lightgrey")
plot(wm_d62, coords, add = TRUE, pch = 19, cex = 0.6, main = "Distance link")
```

### Computing adaptive distance weight matrix

```{r}
knn6 <- knn2nb(knearneigh(coords, k = 6))
knn6
```

```{r}
str(knn6)
```

#### Plotting distance-based neighbours

```{r}
plot(hunan$geometry, border = "lightgrey")
plot(knn6, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

## Weights based on IDW

```{r}
dist <- nbdists(wm_q, coords, longlat = TRUE)
ids <- lapply(dist, function(x) 1/(x))
ids
```

### Row-standardized weights matrix

```{r}
rswm_q <- nb2listw(wm_q, style = "W", zero.policy = TRUE)
rswm_q
```

```{r}
rswm_q$weights[10]
```

```{r}
rswm_ids <- nb2listw(wm_q, glist = ids, style = "B", zero.policy = TRUE)
rswm_ids
```

```{r}
rswm_ids$weights[1]
```

```{r}
summary(unlist(rswm_ids$weights))
```

## Application of Spatial Weight Matrix

### Spatial lag with row-standardized weights

```{r}
GDPPC.lag <- lag.listw(rswm_q, hunan$GDPPC)
GDPPC.lag
```

```{r}
neighbours_1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[neighbours_1]
nb1
```

```{r}
nb1.lag <- GDPPC.lag[neighbours_1]
nb1.lag
```

Comparing the original GDPPC of Polygon 1's five neighbours and those same neighbours' spatially lagged GDPPC values, we observe a difference contributed by the row-standardized weights. Some GDPPC values are increased because of higher contributions by their neighbours and higher weights (fewer neighbours), while some are diluted by having more neighbours with lower GDP.

```{r}
lag.list <- list(hunan$NAME_3, lag.listw(rswm_q, hunan$GDPPC))
lag.res <- as.data.frame(lag.list)
colnames(lag.res) <- c("NAME_3", "lag GDPPC")
hunan <- left_join(hunan, lag.res)
```

```{r}
head(hunan)
```

```{r}
gdppc <- qtm(hunan, "GDPPC")
lag_gdppc <- qtm(hunan, "lag GDPPC")
tmap_arrange(gdppc, lag_gdppc, asp = 1, ncol = 2)
```

### Spatial lag as a sum of neighbouring values

```{r}
b_weights <- lapply(wm_q, function(x) 0*x + 1)
b_weights2 <- nb2listw(wm_q, 
                       glist = b_weights, 
                       style = "B")
b_weights2
```

```{r}
lag_sum <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
lag.res <- as.data.frame(lag_sum)
colnames(lag.res) <- c("NAME_3", "lag_sum GDPPC")
```

```{r}
lag_sum
```

```{r}
hunan <- left_join(hunan, lag.res)
```

```{r}
gdppc <- qtm(hunan, "GDPPC")
lag_sum_gdppc <- qtm(hunan, "lag_sum GDPPC")
tmap_arrange(gdppc, lag_sum_gdppc, asp = 1, ncol = 2)
```

### Spatial window average

```{r}
wm_q1 <- wm_q
```

```{r}
include.self(wm_q1)
```

```{r}
wm_q1 <- nb2listw(wm_q1)
wm_q1
```

```{r}
lag_w_avg_gpdpc <- lag.listw(wm_q1, hunan$GDPPC)
lag_w_avg_gpdpc
```

```{r}
lag.list.wm_q1 <- list(hunan$NAME_3, lag.listw(wm_q1, hunan$GDPPC))
lag_wm_q1.res <- as.data.frame(lag.list.wm_q1)
colnames(lag_wm_q1.res) <- c("NAME_3", "lag_window_avg GDPPC")
```

```{r}
hunan <- left_join(hunan, lag_wm_q1.res)
```

```{r}
gdppc <- qtm(hunan, "GDPPC")
w_avg_gdppc <- qtm(hunan, "lag_window_avg GDPPC")
tmap_arrange(gdppc, w_avg_gdppc, asp = 1, ncol = 2)
```

### Spatial window sum

```{r}
wm_q1 <- wm_q
```

```{r}
include.self(wm_q1)
```

```{r}
wm_q1
```

```{r}
b_weights <- lapply(wm_q1, function(x) 0*x + 1)
b_weights[1]
```

```{r}
b_weights2 <- nb2listw(wm_q1, 
                       glist = b_weights, 
                       style = "B")
b_weights2
```

```{r}
w_sum_gdppc <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
w_sum_gdppc
```

```{r}
w_sum_gdppc.res <- as.data.frame(w_sum_gdppc)
colnames(w_sum_gdppc.res) <- c("NAME_3", "w_sum GDPPC")
```

```{r}
hunan <- left_join(hunan, w_sum_gdppc.res)
```

```{r}
gdppc <- qtm(hunan, "GDPPC")
w_sum_gdppc <- qtm(hunan, "w_sum GDPPC")
tmap_arrange(gdppc, w_sum_gdppc, asp = 1, ncol = 2)
```
